---
title: "The difference between predator-size to prey-size" 
author: Jacob Passfield
output: 
  pdf_document:
    latex_engine: "xelatex"
    number_section: true
    toc: false
    fig_height: 3
    fig_width: 3
    highlight: "tango"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, autodep = TRUE, cache.comments                       = TRUE, fig.align = "center", out.width = "100%")
```

# Predators of Atlantic mackerel and shrimp

The aim of this document is to download a dataset and make some graphs based on it. After searching on <https://www.cefas.co.uk/data-and-publications/fish-stomach-records/> I came across two different species of prey with enough data to make some graphs. The two I chose were the Atlantic mackerel (Scomber scombrus) and caridean shrimps (Caridea) and so I downloaded the two `csv` files containing any record of these two prey being found in the stomach contents of their predators.

## Preparing the data

First of all I load the following packages.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
```

Then I read the data into `R` using the `read_csv()` function.

```{r, message=FALSE}
rawamack <- read.csv("_data/DapstomExportAMack.csv")
rawshrimps <- read_csv("_data/DapstomExportShrimps.csv")
```

After inspecting the data I decided I may be able to create something from the length of the predators and prey recorded. Because of this decision I used the `select()` function to create two new and refined tibbles including the variables of interest. Using the `rename()` function to change the variable names leads to easier coding.

```{r}
#creating a new tibble for Atlantic mackerel
amack <- rawamack %>% 
  rename("Predator" = "Predator.common.name", "PredatorMeanLength" = 
         "Mean.length.of.predator", "Prey" = "Prey.common.name", 
         "PreyLength" = "Prey.Length") %>% 
  select(Predator, PredatorMeanLength, Prey, PreyLength)

#doing the same for caridean shrimp
shrimps <- rawshrimps %>%
  rename("Predator" = "Predator common name", "PredatorMeanLength" = 
           "Mean length of predator", "Prey" = "Prey common name", 
         "PreyLength" = "Prey Length") %>% 
  select(Predator, PredatorMeanLength, Prey, PreyLength) 
```

After experimentation I identified that the data types between the two tibbles did not match and so I ensured they did.

\newpage

```{r, warning=FALSE}
#checking whether data types match
sapply(shrimps, class)
sapply(amack, class)

#changing the data types
amack$Predator <- as.character(as.factor(amack$Predator))
amack$Prey <- as.character(as.factor(amack$Prey))
shrimps$PreyLength <- as.numeric(as.character(shrimps$PreyLength, na.rm=TRUE))
#and now they match
sapply(shrimps, class)
sapply(amack, class)
```

After this I checked the data for any errors or inconsistencies. First for Atlantic mackerel, then for 

```{r}
#changing the capital letters in the predators and prey columns to lowercase letters
amack <- amack %>%
  mutate(Predator = tolower(Predator), Prey = tolower(Prey))
#looking at the range and distribution of the numeric variables
summary(amack$PredatorMeanLength)
summary(amack$PreyLength)
```

`PreyLength` looks okay but the value 587.00 for `PreadtorMeanLength` looks a bit out of place.

\newpage

```{r}
#now visualising the data to idenitfy errors
plot(amack$PredatorMeanLength)
plot(amack$PreyLength)
```

For the sake of this task, I decided to remove this value.

```{r}
#removing the error
amack <- amack %>%
  filter(PredatorMeanLength < 100)
plot(amack$PredatorMeanLength)
```

```{r}
#doing the same with the caridean shrimps

shrimps <- shrimps %>%
  mutate(Predator = tolower(Predator), Prey = tolower(Prey))
summary(shrimps$PredatorMeanLength)
summary(shrimps$PreyLength)
plot(shrimps$PredatorMeanLength)
plot(shrimps$PreyLength)
#removing the abnormally high values
shrimps <- shrimps %>%
  filter(PreyLength < 8)
plot(shrimps$PreyLength)
```

I then decided to combine the two data frames into one tibble.

```{r}
prey <- bind_rows(amack, shrimps)
```

## Basic statistics

I thought of making a bar chart to show the difference between predators for the particular prey chosen.

To do this I created a new tibble which grouped all the data by predator whilst taking the mean length of all combined entries.

```{r}
GroupedPredatorMeanLength <- prey %>% 
  #grouping data by predator and prey
  group_by(Predator, Prey) %>% 
  #summarising the total mean length of predator across all data
  summarise(TotalPredatorMeanLength = mean(PredatorMeanLength,na.rm=TRUE))
```

I then used this to create my box plot.

```{r, fig.width=10, fig.height=10}
barchart1 <- ggplot(data = GroupedPredatorMeanLength, 
                   aes(x = Predator, y = TotalPredatorMeanLength)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Prey) +
  coord_flip()
barchart1
```

Here I realised this bar chart doesn't really compare predator-size with prey-size. What we can see however is that makerel has many more predators than shrimp.

I tried to add a line showing the mean length of the prey using `abline()` but it didn't show up.

At this point I remembered that boxplots are a great comparison tool and so I made some.

```{r}
preyboxplot1 <- ggplot(data = prey, 
                   aes(x=Prey, y=PreyLength)) +
  geom_boxplot()
preyboxplot1
```

Seeing the shrimp boxplot too small to distinguish any value, I used `scale_y_log10()`. 

```{r}
preyboxplot2 <- ggplot(data = prey, 
                   aes(x=Prey, y=PreyLength)) +
  geom_boxplot() +
  scale_y_log10()
preyboxplot2
```

It is clear that mackerel are on average bigger than shrimp.

The next boxplot I made was to comparing the predator length with my two prey.

```{r}
predatorboxplot <- ggplot(data = prey, 
                   aes(x=Prey, y=PredatorMeanLength)) +
  geom_boxplot()
predatorboxplot
```

Here it is evident that mackerel usually have bigger predators compared to shrimp but it is also clear that they can get eaten by the same size predators on occassion.

# Questions/queries I have picked up along the way

1. Should I be doing the optional assignments?
+ 1a. Will our final report be written in RMarkdown?
2. I had trouble understanding what the data was showing, for example, the units and the variables.
3. How do I convert a `.txt` file to `csv`? In reponse to *Barnes et al*.
4. I had this error: your branch is ahead of 'origin/main' by 4 commits. I resolved the issue but I wonder what could've caused it.

Checking to see if I still remember how to push to GitHub.
