---
title: "Initial analysis on how temperature affects fish body sizes #1"
author: Jacob Passfield
output: 
  pdf_document:
    latex_engine: "xelatex"
    number_section: true
    toc: false
    fig_height: 5
    fig_width: 5
    highlight: "tango"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, cache=TRUE, autodep=TRUE, cache.comments=TRUE, fig.align="center", out.width="100%")
```

The following report uses the ideas and data from a report found at *https://www.nature.com/articles/s41559-020-1171-0*. It follows the example given at *https://ourcodingclub.github.io/tutorials/mixed-models/*.

```{r Loading libraries, warning=FALSE, message=FALSE, eval= T}
library(tidyverse)
```

# Exploring the data

```{r Loading data, warning=FALSE, message=FALSE, eval=TRUE , cache.lazy=FALSE}
load(file = "_data/fish_data.RData")
```

I investigate how temperature affects the size of different fish species.

There are 335 species within this dataset. For initial analysis I focused on the fish Trachurus novaezelandiae, commonly known as the yellowtail horse mackerel. This was decided from a quick inspection of the data frame `speciesCount` and finding that yellowtail horse mackerel had the most apperances in the main data file.

```{r Checking data}
# Checking that there are 335 species
length(unique(main_data$TAXONOMIC_NAME))
# Creating a data frame to check how many occurances each species appear in the main data 
# file
speciesCount = as.data.frame(table(main_data$TAXONOMIC_NAME))
```

Created a data frame with observations from yellowtail horse mackerel only.

```{r Creating frame, warning=FALSE, message=FALSE, eval=TRUE, cache.lazy=FALSE}
YelHorMac <- main_data %>% filter(TAXONOMIC_NAME %in% "Trachurus novaezelandiae")
```

# Creating a simple linear regression model

Fitting a linear model is a first step when analysing data.

Size-classes of total fish length (from snout to tip of tail, or longest distance, including for stingrays) used are 2.5, 5.0, 7.5, 10.0, 12.5, 15.0, 20.0, 25.0, 30.0, 35.0, 40.0, 50.0, 62.5 cm, and above. Lengths of fish larger than 62.5 cm should be estimated to the nearest 12.5 cm and individually recorded.

`SizeClass` is the response and `meansst` is the explanatory variable.

```{r Creating basic lm}
basic.lm <- lm(SizeClass ~ meansst, data = YelHorMac)
# lm(formula, data)
# formula here is reponse ~ predictor
# ~ seperates the left- and right-hand sides in a model formula
# the left-hand side is optional and one-sided formulae are used in some contexts
summary(basic.lm)
```

Plotting the model.

```{r Plotting basic lm}
(prelim_plot <- ggplot(YelHorMac, aes(x=meansst, y=SizeClass)) 
   + geom_point() + geom_smooth(method="lm"))
# () around the whole ggplot code creates and shows the graph without the added step of 
# typing prelim_plot after
# method="lm" fits a linear model
# geom_smooth() uses formula y ~ x
# geom_smooth() adds a regression line
# the grey zone is the 95% confidence level interval for predictions from "lm"
```

The higher the mean sea surface temperature, the smaller the size of the yellowtail horse mackerel.

The size class values are distinct, is there anything that we need to do?

Are the assumptions met?

Plotting the residuals should yield a red line that is nearly flat.

```{r}
plot(basic.lm, which=1)
# which=1: a plot of residuals against fitted values
# line not perfectly straight but something they go along with in this example
# for own data be careful
# the bigger the sample size, the less of a trend expected
```

For a Q-Q plot the points should ideally fall along the diagonal dashed line.

```{r}
plot(basic.lm, which=2)
# which=2: a normal Q-Q plot
# points off line
# not good
```

# Transforming data to hopefully lead to better modelling

```{r}
hist(YelHorMac$SizeClass)
```

Histogram with gaps.
In the report, authors did a log-transformation on body length.

```{r}
YelHorMac <- mutate(YelHorMac, logSizeClass = log(SizeClass))
hist(YelHorMac$logSizeClass)
```

Standardising explantory variables (`meansst`) to have a mean of zero (centering) and standard deviation of one (scaling) ensures the estimated coefficients are all on the same scale so it is easier to compare effect sizes.

```{r}
YelHorMac$scaledMSST <- scale(YelHorMac$meansst, center=T, scale=T)
```

`scale()` centers the data (column mean subtracted from the values in the column) and scales it (centered column values divided by the column's standard deviation).

`logSizeClass` is the response and `scaledMSST` is the response.

```{r Creating lm2}
basic.lm2 <- lm(logSizeClass ~ scaledMSST, data = YelHorMac)
# lm(formula, data)
# formula here is reponse ~ predictor
# ~ seperates the left- and right-hand sides in a model formula
# the left-hand side is optional and one-sided formulae are used in some contexts
summary(basic.lm2)
```

Plotting the model.

```{r Plotting basic lm2}
(prelim_plot <- ggplot(YelHorMac, aes(x=scaledMSST, y=logSizeClass)) 
   + geom_point() + geom_smooth(method="lm"))
# () around the whole ggplot code creates and shows the graph without the added step of 
# typing prelim_plot after
# method="lm" fits a linear model
# geom_smooth() uses formula y ~ x
# geom_smooth() adds a regression line
# the grey zone is the 95% confidence level interval for predictions from "lm"
```

The higher the mean sea surface temperature, the smaller the size of the yellowtail horse mackerel.

Plotting the residuals should yield a red line that is nearly flat.

```{r}
plot(basic.lm2, which=1)
# which=1: a plot of residuals against fitted values
# line not perfectly straight but something they go along with in this example
# for own data be careful
# the bigger the sample size, the less of a trend expected
```

For a Q-Q plot the points should ideally fall along the diagonal dashed line.

```{r}
plot(basic.lm2, which=2)
# which=2: a normal Q-Q plot
# points off line
# not good
```

Is our data independent?

```{r}
# Check how many different georgroups there are
length(unique(YelHorMac$geogroup))
```

Since the data collected multiple samples from 43 geogroups, it's expected that data from within each geogroup are more similar to each other than to the data from different georgoup, meaning the observations are correlated.

```{r}
# checking to see if data is correlated
boxplot(SizeClass ~ geogroup, data = YelHorMac)
```

Plotting and colouring points by mountain range.

```{r}
(colour_plot <- ggplot(YelHorMac, aes(x=scaledMSST, y=logSizeClass, colour=geogroup)) +
    geom_point(size=2) + theme_classic() + theme(legend.position="none"))
```

These two plots show that geogroup vary both in temperature and the corresponding body length. This confirms that our observations from within each of the ranges aren't independent. Ignoring this fact could lead to inaccurate conculsions and so we continue.

We could run many seperate analyses and fit a regression for each of the mountain ranges.

```{r}
# data split by mountain range
(split_plot <- ggplot(aes(scaledMSST, logSizeClass), data=YelHorMac) + geom_point() +
    facet_wrap(~geogroup) + xlab("scaled mean sea surface temperature") + ylab("log transformed size class"))
```

As you can see the obstacles we face when running multiple analyses, it becomes rather cluttered and messy.

# Modifying the current model

We want to use all the data in one, but account for data coming from different geogroups.

```{r}
# adding geogroup as a fixed effect to our basic.lm
geo.lm <- lm(logSizeClass ~ scaledMSST + geogroup, data=YelHorMac)
summary(geo.lm)
```

Now body length is not significant. The model above is estimating the difference in log transformed size class between geogroups. But we are not quantifying size for each specific geogroup: want to know whether temeprature affects body length and so we want to control the variation coming from geogroups or, the random effects coming from geogroups. Time for mixed effects models.

# Mixed effects models

A mixed model allows us to use all the data (higher smaple size) and account for the correlations between data coming from the sites and mountain ranges. Also, we'll estimate fewer parameters and avoid porblems with multiple comparisons that we'd encounter while using seperate regressions.

```{r, message=FALSE, warning=FALSE}
library(lme4)
```

Golden rule: want your random effects to have at least five levels. So if we wanted to control the effects of a dragon's sex on intelligence, we would fit sex as a fixed effect, not random (sex: male or female, a two level factor).

Estimating on few data points is very imprecise and so even though you could, there wouldn't be a lot confidence in it.

Randomness doesn't have much to do with mathematical randomness, but more to do with the grouping of variables. It's about making our models representative of our questions and getting better estimates.

The response variable is `logSizeClass` and we are attempting explain part of the variation in body length through fitting mean sea surface temperature as a fixed effect. However the response variable has some residual (unexplained) variation associated with geogroups. We model that unexplained variation through variance by using random effects.

Research question changes slightly to: is there an association between temperature and yellow horse mackerel's body length *after* controlling for the variation in geogroups?

```{r}
# fit the random effect using (1|variableName)
mixed.lmer <- lmer(logSizeClass ~ scaledMSST + (1|geogroup), data=YelHorMac)
summary(mixed.lmer)
```

Need to investigate what this means.

```{r}
plot(mixed.lmer)
# looks alright, no patterns evident
```

Still a decrease? 

```{r}
qqnorm(resid(mixed.lmer))
qqline(resid(mixed.lmer))
# points fall nicely onto the line
```

Points fall on line better. Still off at the end.

Confused :)



